Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-11-26 doc/ApTranscriptome_manuscript.Rmd
updates to files for submission to Molecular Ecology

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

# Significance Statement

Coping with progressive changes in climate requires both acclimation and long-term adaptation. To better understand the relationship between these mechanisms, we compared gene expression as a function of temperature in two ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. Aphaenogaster picea responded to extreme temperatures primarily through upregulation, with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a shift in mechanisms from those used to cope with rarely encountered stressors.

@@

@@

@@

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

@@

# Material and Methods

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supplemental Methods 1) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods 1).

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

## Data availability

Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626. The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at [TBD: Harvard Forest Archives]. Reproducible R script for analysis is available from GitHub [@stanton-geddes2014].

## Acknowledgements

@@

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

@@

# Materials and Methods

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Methods). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supplemental Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods).

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental Methods).

@@

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

@@

# Acknowledgements

@@

## Data Accessibility

Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626. The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive). The reproducible and version-controlled script for data analysis is available from GitHub (https://github.com/johnstantongeddes/ApTranscriptome).

## Author Contributions

JSG, NG and SHC designed research. JSG, ADN and LC performed research. JSG, JV, MV and SCH analyzed data and wrote the paper. ADN and LC performed research. JV and MV analyzed data. JSG, RD, AE, NS, NG and SHC wrote the paper.

### 2014-11-26 doc/MolEco_cover_letter.tex
updates to files for submission to Molecular Ecology

\noindent{Dear \emph{PNAS} Editorial Board,}

We are excited to submit our manuscript \emph{Thermal
reactionomes reveal adaptive responses to thermal extremes in warm and
cold-climate ant species} for consideration as a Research Report in
\emph{The Proceedings of the National Academy of Sciences}. There are
several aspects of the work that we think make it worthy of publication
in \emph{PNAS}.

@@

This manuscript is most appropriate for the major classification
BIOLOGICAL SCIENCES, minor classification Ecology.

\textbf{Suggested Editors:} Douglas J. Futuyma, David L. Denlinger, David M. Hillis

\textbf{Suggested NAS Members:} Trudy F. Mackay, Nevo Eviata, George N. Somero

\textbf{Suggested reviewers:}  Dr. Nicholas Teets, Postdoc, University of Florida; Dr. Vanessa Kellerman, Postdoc, Monash University; Dr. Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies; Dr. Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University; Dr. Joel Kingsolver, Professor, University of North Carolina

>>>>

\noindent{Dear Dr. Riesberg,}

We are excited to submit our manuscript \emph{Thermal
reactionomes reveal adaptive responses to thermal extremes in warm and
cold-climate ant species} for consideration as an Original Article in
\emph{Molecular Ecology}. There are
several aspects of the work that we think make it worthy of publication
in \emph{Molecular Ecology}.

@@

\textbf{Suggested reviewers:}

\begin{itemize}
  \item Dr. Nicholas Teets - expertise with insect genomics and transcriptomics
  \item Dr. Vanessa Kellerman - expertise with insect population and quantitative genetics
  \item Dr. Pavlos Pavlidis - critical insight on function annotation and gene ontologies
  \item Dr. Joel Kingsolver - expertise in insect thermal limits
  \item Dr. Daniel Barshis - expertise in ecological genomics
\end{itemize}

\textbf{Suggested editors:}

\begin{itemize}
  \item Dr. Ary Hoffman - expertise with insect population and quantitative genetics
  \item Dr. Stephen Palumbi - expertise in ecological genomics
  \item Dr. Aurelie Bonin - expertise in population genomics
\end{itemize}
